Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-10 paper.tex

@@

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
% 
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm 
to simulate the coalescent with
recombination was formulated backwards in time
by Hudson \citeyearpar{hudson_properties_1983}. 
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process 
\citep{wiuf_recombination_1999}.
Such an along-the-genome approach is considerably more complex as the next 
tree to be simulated depends on all previous trees.
This idea however provided the basis of the 
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time, 
as well as along-the-genome.
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} 
genomic intervals may coalesce. (The SMC' additionally allows 
lineages with abutting genomic intervals to coalesce.)
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.
Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of
the ARG, the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, this is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe
the ancestral
material associated with each common ancestor node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
which each are given new, unique labels and added to $L(t)$
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Calculating likelihoods} \label{par:liks}
[Paragraph setting the scene on likelihood calculations]
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present
[Also: \cite{guo_recombination-aware_2022}]
[NOTE the following is super weak, and needs a lot of work.]

>>>>

[Things to include: locally unary nodes rather than explicit events.
edge-centric formulation vs event centric. Graceful degradation of
performance with incomplete ARGs vs complete inability to compute.]

@@

% coalescent introduction: backwards-in-time
% left-to-right: Wiuf-Hein (difficult)
% SMC: although easy to reason about left-to-right sense
% 
The coalescent process describes the ancestry of a set of sampled genomes.
The original algorithm 
to simulate the coalescent with
recombination was formulated backwards in time
by Hudson \citeyearpar{hudson_properties_1983}. 
Wiuf and Hein later described
a tree-by-tree method to simulate the same stochastic process 
\citep{wiuf_recombination_1999}.
Such an along-the-genome approach is considerably more complex as the next 
tree to be simulated depends on all previous trees.
This idea however provided the basis of the 
SMC \citep{mcvean_approximating_2005}, restricting the set of all state
space transitions possible under the Hudson coalescent to
obtain a process that is Markovian both backwards-in-time, 
as well as along-the-genome.
In its backwards-in-time formulation, the SMC only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} 
genomic intervals may coalesce. The SMC' is a refinement 
of the SMC which also allows lineages with abutting genomic intervals to 
coalesce~\citep{marjoram_fast_2006}.
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,
rasmussen_genome-wide_2014}.

Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of ARGs, 
the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, that is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
We require a single half-open interval to describe
the ancestral
material associated with each common ancestor node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards in time, $t=0$ is today, and 
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
given new, unique labels and added to $L(t)$
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Likelihoods} \label{par:liks}

Computing the likelihood of a putative ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner, Yamato \& Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
decomposes the likelihood of an ARG under the coalescent with recombination
into a product of the probablility of 
each common ancestor event, each recombination event and the probablility of
the inter-event waiting times. In essence, computing the KYF likelihood requires
``replaying'' the full state of Hudson's simulation
algorithm~\citep{hudson_properties_1983,kelleher_efficient_2016} in order to
compute the probability of the configuration of lineages and distribution
of ancestral material at each time point. It therefore requires complete
detail about each of these events to be stored. 
Such complete detail is not present in many estimated
ARGs~\citep{wong_general_2023} or even in most simulations.
Indeed, supporting likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to the \texttt{msprime} simulator~\citep{baumdicker_efficient_2021}.
More recent approaches to defining likelihoods on such models require
similarly exhaustive input\citep[e.g.][]{guo_recombination-aware_2022} (MORE?
Anything for GC in bacteria, etc?)
